A Formal Model of Checked C

### 2023-01-18 sandp23/introduction.tex
update introduction.

Vulnerabilities due to memory corruption, especially spatial memory corruption, 
are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe, Zeng:2013:SRF:2534766.2534798} 
despite a large body of work that tries to prevent them~\cite{song2019sanitizing}.
Several industrial and research efforts, including CCured~\cite{Necula2005},
Softbound~\cite{softbound}, and ASAN~\cite{Serebryany2012},
have investigated ways to
\mzu{compile C programs 
  to automatic spatial safety enforcement.}
\mz{You cannot compile a program to enforcement.}
% 
These approaches all impose performance overheads deemed too high for deployment use. 
% 
\liyi{Should we say this? Is the paper about overhead reducing? }
Recently,~\citet{Elliott2018} and~\citet{li22checkedc} introduced and formalized \checkedc, an
open-source extension to C
% 
\mzs{ with new types and annotations to ensure a program’s spatial safety.
  Specifically, \checkedc extends C with checked pointer types, which are
  restricted by the compiler to spatially safe uses.}
% 
\mzr{to ensure a program’s spatial safety by introducing additional annotations and
  enforcing new types, checked pointer types in particular.}
% 
\mz{Redundancy?}
% 
The checked pointers are represented as system-level memory words\mzs{,~\ie} without ``fattening'' metadata, ensuring backward compatibility.
\mz{\ie{} sounds indirect and redundant to me.}
% 
\mzu{This enables incremental conversion, where regions of code can be converted to~\checkedc and ensure spatial safety within those regions.}
\mz{This sentence is incoherent. I don't understand the connection between
  incremental conversion and regions of code. 
  % 
  I guess your point is that since you don't introduce extra metadata, 
  checked and unchecked region can coexist and interact seamlessly.
  % 
}
% 
However, as we explain in~\sect{subsec:nosafetyagsintuncheckedcode}, the
unconverted or unchecked code can violate guarantees provided in \mzu{Checked}
\mz{either tt it or use lowercase?}
regions.
%
We need to ensure that~\emph{code executed as part of unchecked regions does not violate the safety guarantees in checked regions}.

% One possible solution is Checked C
% \begin{itemize}
% \item Think of it as migratory/gradual typing, but for C.
% \item Distinct pointer types, but which are backward binary- and source-compatible.
% \item Checked regions, containing only checked pointers and restricted
%   idioms, aim to ensure spatial safety
% \item Implemented as an extension to Clang/LLVM. Good performance
%   (compared to ASAN etc.)
% \end{itemize}
\iffalse
\noindent
\textbf{No safety against unchecked code.} The \checkedc spatial safety guarantee applies to completely converted programs, i.e., programs that uses only checked types.
Although, the backward compatibility of Checked C helps a partially annotated program to enjoy spatial memory safety on those regions using only Checked pointers (i.e., checked or safe regions).
But, the unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}.
Although the blameless proof exists~\cite{ruef2019achieving} for~\checkedc, it does not state that spatial safety violations cannot happen in Checked regions but rather states that Checked regions~\emph{cannot be blamed for any spatial safety violations}.
Consider the following example:
\begin{minted}[xleftmargin=25pt, mathescape, linenos, escapeinside=||, fontsize=\footnotesize{}]{c}
// Checked code
int func(array_ptr<char> p : count(5)) {
|\textcolor{red}{\faChainBroken}|..p[4]..
}
// unchecked code
...
str = "he";
...
|\textcolor{red}{\faBug}|assume_bounds_cast<char>(str, 5); 
...
char ptr[16];
...
len <- derived from user input
...
|\textcolor{red}{\faBug}| memcpy(ptr, buff, len); // buffer overflow
\end{minted}
Here, the checked function~\inlinecode{func} expected a pointer to a buffer of five elements, but unchecked code violated it and invoked the function with a buffer of 2 elements.
This results in a spatial safety violation (\textcolor{red}{\faChainBroken}) in the Checked region, but of course, the blame or the root cause is in the unchecked region (\textcolor{red}{\faBug}).
Furthermore, since checked and unchecked regions execute in the same address space, spatial memory corruptions in unchecked regions (Line 15) can take down the complete program despite having checked regions.
We need~\emph{an isolation mechanism to ensure that code executed as part of unchecked regions does not violate the safety guarantees in checked regions}.
\fi
% 
This mechanism is called program partitioning~\cite{rul2009towards}.
Existing mechanisms are not suitable as they are based on process isolation and have high overhead, and are data-centric~\emph{data-centric} (\sect{subsec:background:programpart}).
Furthermore, these techniques are hard to engineer to co-exist with~\checkedc.
\liyi{why? Can we not say this? It seems that the whole reason people should care our work is we did it in \checkedc, but \checkedc itself is the selling product here.... }
But in our case, we want a low-overhead code-centric partitioning, where the unchecked code (or functions) should be isolated (or partitioned) from checked code. We also want the technique to co-exist and be compatible with Checked C guarantees such that the partition containing checked code should still enjoy its spatial safety.

In this work, we propose a type-directed code-centric program partitioning approach.
Specifically, our system,~\systemname, extends Checked c using~\textbf{tainted} (\taintt) types, which can be used to mark functions and pointers that need to be isolated from the original program.
The checked types follow the standard Checked C typing rules.
However, our type-system disallows any unsafe interactions between taint and untainted types.
% 
\mz{It sounds hard to follow. The subjects in these sentences are  ``our
  system''. ``checked types'' and ``our type-system''.
  %
  For readers, the connection among checked, tainted and unchecked seems weak.
  % 
} 
% 
% Specifically, in our system,~\systemname, extending CheckedC,
% functions and pointers that need to be isolated from the original program can be
% marked with ~\textbf{tainted} (\taintt) types.
%

% 
The developer starts by marking desired (\ie unchecked) functions and pointers used in these functions as tainted.
Second,~\systemname partitions the given program into two partitions (\umode and~\cmode regions)  of different \mzu{privileges}:
% 
\mz{need a bit more description (4-5 words) before introducing the word
  ``privilege''}
%

\begin{itemize}
\item~\umode \emph{region} or~\emph{tainted region} (low privilege tainted region): This partition contains tainted types (\ie functions and pointers) and can only access tainted pointers.
\item~\cmode \emph{region} or~\emph{safe region} (high privilege untainted or checked region): This partition contains the remaining (untainted) code and data and has complete access to~\cmode region.
The functions in~\cregion can invoke any function in~\ucregion and access all its data but not the other way around, except for call-back functions, which we will discuss later.
\end{itemize}
% 
\mz{                            %
  This paragraph is unclear. 
  % 
  How does regions look like? Can one region nested in another?
  %
  If not, ``This partition contains the remaining (untainted) code and data and
  has complete access to~\cmode region.'' creates a sense of recursive
  definition of \cregion.
}
%

% 
Finally, during program execution, the~\ucregion partition will be executed in an existing sandboxed environment (\eg WASM sandbox), and our compiler will add the necessary instrumentation to facilitate the communication between code in~\cregion{} and~\ucregion{}.

The combination of tainted types and privileged partitions enables us to enforce isolation and provide memory safety without marshaling costs.
% 
\mzu{As functions in the~\ucregion can only access tainted types (or unchecked
types),~\cregion functions should use tainted types to pass pointer arguments
to~\ucregion functions.
% 
We avoid marshaling by allocating all tainted pointers (\ie tainted buffers)
in~\ucregion and thus can be accessed in both partitions.}
%
\mz{I don't quite understand the logical consequence here. Why does allocating
  tainted points avoids marchaling. 
  % 
  I see that ``allocating all tainted pointers'' makes ptrs accessable in both
  regions, but avoiding marchaling doesn't seem to cause that.
  %
  The question I should probably ask is what marshaling means? 
  % 
  Does the procedure that copies a checked buffer to a tainted one counts as
  marshaling?
  % 
}
% 
Although memory isolation prevents direct violations,~\ucregion code can still affect~\cregion through tainted pointers by confused deputy attacks.
Our compiler avoids these attacks by ensuring, through dynamic checks that tainted pointers validly point to \ucregion address space.
Such checks are statically generated by our compiler.

@@

\myparagraph{\systemname Type System, Formalism and Compiler}
We present a type system that integrates tainted types with~\checkedc and ensures the~\emph{non-crashing property},~\ie 
a well-typed~\systemname program can never crash due to spatial safety violations of tainted pointers.
We extend~\checkedc compiler to support our type system.
We formalize our system by extending~\checkedc formalism~\citet{li22checkedc} with the non-crashing guarantee and other new features below.
We formally prove the~\emph{non-crashing theorem} and use model-based randomized testing (CITE) to certify the simulation relation between the~\systemname semantics and the compiler.
To the best of our knowledge,~\systemname is the first C(-like) language and compiler formalism with the program partitioning mechanism.
%
\mz{uh}

\myparagraph{Type-Directed Program Partitioning}
We present a type-directed program partition technique to separate checked and unchecked code regions and ensure that the code in checked region~\textit{cannot crash} due to spatial safety violations.
Our type system restricts checked pointers usage only in~\cregion and ensures that~\ucregion can only access tainted pointers.
Our modular design enables us to use existing sandboxed techniques to enforce memory and execution isolation.

\myparagraph{Supporting callbacks to~\cregion with no Checked Pointer Exposure}
Although we disallow access to~\cregion from~\ucregion, there can be cases where such access is needed.
Specifically, when~\cregion wants to provide access to certain shared checked data to~\ucregion.
To enable this, we support callback functions in~\cregion that can be invoked from~\ucregion through function pointers.
%However, knowing the address of~\cregion functions in~\ucregion violates the program partition principle of \systemname.
However, knowing the address of~\cregion functions in~\ucregion violates the program partition principle of~\systemname and leads to other attacks~\cite{hauser2019sleak}.

We handle this by using indirection. Specifically, instead of directly accessing the~\cregion callbacks, the ~\ucregion accesses them using a protected trampoline function, which will direct the execution to the appropriate callback function.
In addition, the trampoline function itself will be referenced using an opaque index rather than its virtual address, and \mzu{we will utilize existing sandboxing techniques to implement this}.\mz{tense?}
% 
In our \systemname formalism, we formally verified that our formalism is
\emph{non-exposure}, \ie, no \cregion pointer addresses will be leaked in
\ucregion code.

@@

We evaluated~\systemname by partioning~\numprog{} large real-world programs to demonstrate its effectiveness.
Our evaluation shows that~\systemname provides a flexible, low-overhead program partitioning mechanism and guarantees spatial memory safety.

\aravind{Fix the following text after all the sections of the papers are finalized.}
\liyi{the following might be removed if space is needed, since it is usually not quite useful.}
We begin with a review of \checkedc{} \mzr{and introduction of} new features in
\systemname (Section~\ref{sec:overview}), present our main contributions
(Sections~\ref{sec:formal}--\ref{sec:evaluation}), and conclude with a
discussion of related and future work (Sections~\ref{sec:related},
\ref{sec:conclude}). All code and proof artifacts (both for Coq and Redex) can
be found at \url{https://github.com/plum-umd/checkedc}.

>>>>

Vulnerabilities due to memory corruption, especially spatial memory corruption, 
are still a major issue for C programs~\cite{cvetrend, microsoftmemsafe, Zeng:2013:SRF:2534766.2534798} 
despite many efforts that tried to prevent them~\cite{song2019sanitizing}.
Several industrial and research efforts, including CCured~\cite{Necula2005},
Softbound~\cite{softbound}, and ASAN~\cite{Serebryany2012},
have investigated ways to better compile C programs with automatic spatial safety enforcement.
These approaches all impose performance overheads deemed too high for deployment use. 
% 
%\liyi{Should we say this? Is the paper about overhead reducing? }
Recently,~\citet{Elliott2018} and~\citet{li22checkedc} introduced and formalized \checkedc, an
open-source extension to C,
to ensure a program’s spatial safety by introducing additional annotations and
enforcing new types, checked pointer types in particular.
%
The checked pointers are represented as system-level memory words without ``fattening'' metadata, 
and ensuring backward compatibility, i.e.,
users are able to split code into checked and unchecked regions and incrementally convert
C code in the unchecked region to Checked C code in the checked one.
However, as we explain in~\sect{subsec:nosafetyagsintuncheckedcode}, the
unconverted or unchecked code can violate guarantees provided in checked
regions.
%
We need to ensure that~\emph{code executed as part of 
unchecked regions does not violate the safety guarantees in checked regions  with the use of program partitioning mechanism \cite{rul2009towards}}.

% One possible solution is Checked C
% \begin{itemize}
% \item Think of it as migratory/gradual typing, but for C.
% \item Distinct pointer types, but which are backward binary- and source-compatible.
% \item Checked regions, containing only checked pointers and restricted
%   idioms, aim to ensure spatial safety
% \item Implemented as an extension to Clang/LLVM. Good performance
%   (compared to ASAN etc.)
% \end{itemize}
\iffalse
\noindent
\textbf{No safety against unchecked code.} The \checkedc spatial safety guarantee applies to completely converted programs, i.e., programs that uses only checked types.
Although, the backward compatibility of Checked C helps a partially annotated program to enjoy spatial memory safety on those regions using only Checked pointers (i.e., checked or safe regions).
But, the unconverted code regions (or unsafe regions) can affect pointers in safe regions and violate certain assumptions leading to vulnerabilities, as demonstrated by cross-language attacks~\cite{mergendahlcross}.
Although the blameless proof exists~\cite{ruef2019achieving} for~\checkedc, it does not state that spatial safety violations cannot happen in Checked regions but rather states that Checked regions~\emph{cannot be blamed for any spatial safety violations}.
Consider the following example:
\begin{minted}[xleftmargin=25pt, mathescape, linenos, escapeinside=||, fontsize=\footnotesize{}]{c}
// Checked code
int func(array_ptr<char> p : count(5)) {
|\textcolor{red}{\faChainBroken}|..p[4]..
}
// unchecked code
...
str = "he";
...
|\textcolor{red}{\faBug}|assume_bounds_cast<char>(str, 5); 
...
char ptr[16];
...
len <- derived from user input
...
|\textcolor{red}{\faBug}| memcpy(ptr, buff, len); // buffer overflow
\end{minted}
Here, the checked function~\inlinecode{func} expected a pointer to a buffer of five elements, but unchecked code violated it and invoked the function with a buffer of 2 elements.
This results in a spatial safety violation (\textcolor{red}{\faChainBroken}) in the Checked region, but of course, the blame or the root cause is in the unchecked region (\textcolor{red}{\faBug}).
Furthermore, since checked and unchecked regions execute in the same address space, spatial memory corruptions in unchecked regions (Line 15) can take down the complete program despite having checked regions.
We need~\emph{an isolation mechanism to ensure that code executed as part of unchecked regions does not violate the safety guarantees in checked regions}.
\fi
%

Existing such mechanisms are not suitable as they are based on process isolation and have high overhead, and are \emph{data-centric} (\sect{subsec:background:programpart}).
%Furthermore, these techniques are hard to engineer to co-exist with~\checkedc.
%\liyi{why? Can we not say this? It seems that the whole reason people should care our work is we did it in \checkedc, but \checkedc itself is the selling product here.... }
But in our case, we want a low-overhead code-centric partitioning, where the unchecked code (or functions) should be isolated (or partitioned) from checked code. We also want the technique to co-exist and be compatible with Checked C guarantees such that the partition containing checked code should still enjoy its spatial safety.

Here, we propose a type-directed code-centric program partitioning approach.
Specifically, our system,~\systemname, extends \checkedc's checked and unchecked pointer types---representing safe and unsafe program pieces---with \textbf{tainted} (\taintt) types running on an isolated sandbox mechanism,
forbids the communication between checked and unchecked type entities, and enforces the communication between checked and unchecked types
through the uses of tainted types with additional validity checks. 
%which can be used to mark functions and pointers that need to be isolated from the original program, while checked types follow the standard \checkedc typing rules.
%However, our type-system disallows any unsafe interactions between taint and untainted types.
% 
% Specifically, in our system,~\systemname, extending CheckedC,
% functions and pointers that need to be isolated from the original program can be
% marked with ~\textbf{tainted} (\taintt) types.
%

The developer starts by marking desired (\ie unchecked) functions and pointers used in functions as tainted. 
Then, \systemname partitions the given program into two partitions (\umode and~\cmode regions)  of different privileges:

\begin{itemize}
\item~\umode \emph{region} or~\emph{tainted region} (low privilege tainted region): This partition contains tainted types (\ie functions and pointers) and can only access tainted pointers.
\item~\cmode \emph{region} or~\emph{safe region} (high privilege untainted or checked region): This partition contains the remaining (untainted) code and data and has complete access to~\cmode region.
The functions in~\cregion can invoke any function in~\ucregion and access all its data but not the other way around, except for call-back functions, which we will discuss later.
\end{itemize}

The $\cmode$ region code is executed by compiling the code to a normal \checkedc compiler, while the~\ucregion partition will be executed in an existing sandboxed environment (\eg WASM sandbox), and our compiler adds necessary instrumentations to facilitate the communication between code in~$\cmode$ and~$\umode$ regions.

The combination of tainted types and privileged partitions 
enables us to enforce isolation and provide memory safety without transforming all unchecked C code to \checkedc code,
because unchecked types can stay in \ucregion, and \cregion code can access tainted type entities that are allocated 
in \ucregion.
% 
Although memory isolation prevents direct violations,~\ucregion code can still affect~\cregion through tainted pointers by confused deputy attacks.
Our compiler avoids these attacks by ensuring, through dynamic checks that tainted pointers validly point to \ucregion address space.
Such checks are statically generated by our compiler.

@@

\myparagraph{\systemname Type System, Formalism and Compiler}
We present a type system that integrates tainted types with~\checkedc and ensures the~\emph{non-crashing} and \emph{non-exposure} guarantees,~\ie a well-typed~\systemname program can never crash due to spatial safety violations,
as well as \ucregion code cannot directly observe a checked pointer address.
%no \cregion pointer addresses will be leaked in
%\ucregion code
We extend~\checkedc compiler to support our type system.
We formalize our system by extending~\checkedc formalism~\citet{li22checkedc} with the non-crashing and non-exposure guarantees.
We formally prove theorems related to the two guarantees and use model-based randomized testing (CITE) to certify the simulation relation between the~\systemname semantics and its compiler formalism.
To the best of our knowledge,~\systemname is the first C(-like) language and compiler formalism with the program partitioning mechanism.

\myparagraph{Type-Directed Program Partitioning}
We present a type-directed program partition technique to separate $\cmode$ and $\umode$ code regions and ensure the above guarantees.
%Our type system restricts checked pointers usage only in~\cregion and ensures that~\ucregion can only access tainted pointers.
Our modular design enables us to use existing sandboxed techniques to enforce memory and execution isolation, with the implementation of tainted pointers in the \systemname compiler.

\myparagraph{Supporting callbacks to~\cregion with no Checked Pointer Exposure}
Although we disallow access to~\cregion from~\ucregion, there can be cases where such access is needed.
Specifically, when~\cregion wants to provide access to certain shared checked data to~\ucregion.
To enable this, we support callback functions in~\cregion that can be invoked from~\ucregion through function pointers.
%However, knowing the address of~\cregion functions in~\ucregion violates the program partition principle of \systemname.
However, knowing the address of~\cregion functions in~\ucregion violates the non-exposure guarantee and leads to other attacks~\cite{hauser2019sleak}.

We handle this by using indirection. Specifically, instead of directly accessing the~\cregion callbacks, the ~\ucregion accesses them using a tainted-typed protected trampoline function, which directs the execution to the appropriate callback function.
In addition, the trampoline function itself is referenced using an opaque index rather than its virtual address, implemented through
existing sandboxing techniques.
% 
%In our \systemname formalism, we formally verified that our formalism is
%\emph{non-exposure}, \ie, no \cregion pointer addresses will be leaked in
%\ucregion code.

@@

We evaluated~\systemname by partioning six large real-world programs to demonstrate its effectiveness.
Our evaluation shows that~\systemname provides a flexible, low-overhead program partitioning mechanism and guarantees spatial memory safety.

\ignore{
\aravind{Fix the following text after all the sections of the papers are finalized.}
\liyi{the following might be removed if space is needed, since it is usually not quite useful.}
We begin with a review of \checkedc{} \mzr{and introduction of} new features in
\systemname (Section~\ref{sec:overview}), present our main contributions
(Sections~\ref{sec:formal}--\ref{sec:evaluation}), and conclude with a
discussion of related and future work (Sections~\ref{sec:related},
\ref{sec:conclude}). All code and proof artifacts (both for Coq and Redex) can
be found at \url{https://github.com/plum-umd/checkedc}.
}
